Perturbing SU($n$)-symmetric interactions

### 2020-02-04 collective_sun/sun_perturbation.tex
added text (in progress) formalizing diagrams and explaining rules to decompose/manipulate them

To simplify the product in \eqref{eq:sym_prod_proj}, we first write
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\C_N\p{M_j}} w_j\p{k} O_j\p{k}
  = \prod_{j\in\ZZ_p} \f1{M_j!} \sum_{k\in\ZZ_N^{M_j}}
  w_j\p{k} O_j\p{k},
  \label{eq:sym_prod_vec}
\end{align}
where we extend the definitions of $w_j\p{k}$ and $O_j\p{k}$ to all
vectors $k\in\ZZ_N^{M_j}$ by defining symmetric rank-$M_j$ tensors
$w_j$ and $O_j$ (where $O_j$ is an {\it operator-valued} tensor) for
which $w_j\p{k}=O_j\p{k}=0$ if $k$ contains any repeated symbols.  We
then collect terms whose operator content is equivalent up to a
permutation of spins.  To this end, we classify terms in
\eqref{eq:sym_prod_vec} by the numbers $g_S$ of indices shared by all
tensors $w_j,O_j$ with $j\in S\subset\ZZ_p$.  For example, a term of
the form $w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e}$ with distinct
indices $\p{a,b,c,d,e}\in\C_N\p{5}$ would have
\begin{align}
  g_{\set{1}} = 1,
  &&
  g_{\set{1,2,3}} = 1,
  &&
  g_{\set{1,3}} = 1,
  &&
  g_{\set{2,3}} = 2,
\end{align}
and $g_S=0$ for all other subsets $S\subset\ZZ_3$.  This index
assignment can be represented by the Venn diagram
\begin{align}
  \diagram{example},
  \label{eq:venn_diagram}
\end{align}
where $g_S$ is determined the number of dots at the intersection of
circles $j\in S$.  We denote an assignment of distinct indices
according to choices of $g_S$ for all $S\in\PP\p{\ZZ_p}$ by $g$, where
$\PP\p{\ZZ_p}$ is the power set (i.e.~set of all subsets) of $\ZZ_p$.
Given a vector $\m M$ of the tensor ranks $M_j$, we then denote the
set of all valid index assignments $g$ by $\G\p{\m M}$.  For
consistency, all valid index assignments $g\in\G\p{\m M}$ must assign
exactly $M_j$ indices to the rank-$M_j$ tensor $w_j$, i.e.~
\begin{align}
  \sum_{S\in\PP\p{\ZZ_p}\,:\,j\in S} g_S
  = \sum_{R\in\PP\p{\ZZ_p\setminus\set{j}}} g_{\set{j}\cup R}
  = M_j
\end{align}
for all $j\in\ZZ_p$.  In order to make the index assignments
determined by choices of $g_S$ for all $S\in\PP\p{\ZZ_p}$ unique, we
also enforce that $g_{\set{}}=0$ for all $g\in\G\p{\m M}$, which is
equivalent to saying that every dot in the corresponding Venn diagram
must lie within some circle.  The set of valid index assignments
$\G\p{\m M}$ is then essentially the set of all Venn diagrams of the
form in \eqref{eq:venn_diagram} with $p=\rank\p{\m M}$ circles and a
total of $M_j$ dots within circle $j$.

@@

In order to write out the factors in \eqref{eq:sym_prod_group_start}
explicitly, we identify the set of values that the distinct indices
assigned according to $g$ can take:
\begin{align}
  \C_N\p{g}
  \equiv \set{ k \in \bigotimes_{S\in\PP\p{\ZZ_p}} \C_N\p{g_S}
    : \t{all elements of $k$ are distinct} },
\end{align}
where each $k\in\C_N\p{g}$ has the natural decomposition
$k=\p{k_S:S\in\PP\p{\ZZ_p}}$ with $k_S\in\C_N\p{g_S}$.  We then define
the restriction of $k\in\C_N\p{g}$ to values associated with $w_j$ and
$O_j$:
\begin{align}
  k_j \equiv \bigcup_{S\in\PP\p{\ZZ_p} : j\in S} k_S
  = \bigcup_{R\in\PP\p{\ZZ_p\setminus\set{j}}} k_{\set{j}\cup R},
\end{align}
where the union of lists $k_S$ denotes a concatenation of those lists.
These definitions allow us to expand
\begin{align}
  w\p{g} \equiv \sum_{k\in\C_N\p{g}} \prod_{j\in\ZZ_p} w_j\p{k_j},
  &&
  O\p{g} \equiv \prod_{j\in\ZZ_p}
  O_j\p{\ell_j}~\t{for some}~\ell\in\C_N\p{g},
\end{align}
where the choice of $\ell\in\C_N\p{g}$ does not matter after
projection onto permutationally symmetric manifold, and
\begin{align}
  \S\p{g} \equiv \sp{\prod_{j\in\ZZ_p} { M_j \choose g_S : j\in S }}
  \sp{\prod_{S\in\PP\p{\ZZ_p}} \p{g_S!}^\abs{S}}
  = \prod_{j\in\ZZ_p} M_j!,
  \label{eq:sym_prod_symmetry}
\end{align}
where the first product in \eqref{eq:sym_prod_symmetry} contains
multinomial coefficients
\begin{align}
  { M \choose a_1, a_2, \cdots, a_m }
  \equiv \f{M!}{a_1!a_2!\cdots a_m!},
\end{align}
that account for the number of ways to partition the indices of each
$w_j,O_j$ into sets of shared indices according to $g_S$ for all
$j\in S$, and the second product in \eqref{eq:sym_prod_symmetry}
accounts for the number of ways to permute each set of shared indices
on every tensor.  Altogether, the symmetry factors $\S\p{g}$ cancel
out with the $\sim M_j!$ prefactors in \eqref{eq:sym_prod_vec}, so
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\C_N\p{M_j}} w_j\p{k} O_j\p{k}
  \EQFS \sum_{g\in\G\p{\m M}} w\p{g} O\p{g}.
  \label{eq:sym_prod_group}
\end{align}

@@

>>>>

To simplify the product in \eqref{eq:sym_prod_proj}, we first write
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\C_N\p{M_j}} w_j\p{k} O_j\p{k}
  = \prod_{j\in\ZZ_p} \f1{M_j!} \sum_{k\in\ZZ_N^{M_j}}
  w_j\p{k} O_j\p{k},
  \label{eq:sym_prod_vec}
\end{align}
where we extend the definitions of $w_j\p{k}$ and $O_j\p{k}$ to all
vectors $k\in\ZZ_N^{M_j}$ by defining symmetric dimension-$M_j$
tensors $w_j$ and $O_j$ (i.e.~tensors with $M_j$ indices, with $O_j$
an {\it operator-valued} tensor) for which $w_j\p{k}=O_j\p{k}=0$ if
$k$ contains any repeated symbols.  We then collect terms whose
operator content is equivalent up to a permutation of spins.  To this
end, we classify terms in \eqref{eq:sym_prod_vec} by the numbers $g_S$
of indices shared by all tensors $w_j,O_j$ with $j\in S\subset\ZZ_p$.
For example, a term of the form
$w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e}$ with distinct indices
$\p{a,b,c,d,e}\in\C_N\p{5}$ would have
\begin{align}
  g_{\set{1}} = 1,
  &&
  g_{\set{1,2,3}} = 1,
  &&
  g_{\set{1,3}} = 1,
  &&
  g_{\set{2,3}} = 2,
\end{align}
and $g_S=0$ for all other subsets $S\subset\ZZ_3$.  This index
assignment can be represented by the Venn diagram
\begin{align}
  \diagram{example_triple},
  \label{eq:venn_diagram}
\end{align}
where $g_S$ is determined the number of dots at the intersection of
circles $j\in S$.  We denote an assignment of distinct indices
according to choices of $g_S$ for all $S\in\PP\p{\ZZ_p}$ by $g$, where
$\PP\p{\ZZ_p}$ is the power set (i.e.~set of all subsets) of $\ZZ_p$.
Given a vector $\m M$ of the dimensions $M_j$, we then denote the set
of all valid index assignments $g$ by $\G\p{\m M}$.  For consistency,
all valid index assignments $g\in\G\p{\m M}$ must assign exactly $M_j$
indices to the dimension-$M_j$ tensor $w_j$, i.e.~
\begin{align}
  \sum_{S\in\PP\p{\ZZ_p}\,:\,j\in S} g_S
  = \sum_{R\in\PP\p{\ZZ_p\setminus\set{j}}} g_{\set{j}\cup R}
  = M_j
\end{align}
for all $j\in\ZZ_p$.  In order to make the index assignments
determined by choices of $g_S$ for all $S\in\PP\p{\ZZ_p}$ unique, we
also enforce that $g_{\set{}}=0$ for all $g\in\G\p{\m M}$, which is
equivalent to saying that every dot in the corresponding Venn diagram
must lie within some circle.  The set of valid index assignments
$\G\p{\m M}$ is then essentially the set of all Venn diagrams of the
form in \eqref{eq:venn_diagram} with $p=\dim\p{\m M}$ circles and a
total of $M_j$ dots within circle $j$.

@@

In order to write out the factors in \eqref{eq:sym_prod_group_start}
explicitly, we identify the set of values that the distinct indices
assigned according to $g$ can take:
\begin{align}
  \C_N\p{g}
  \equiv \set{ k \in \bigotimes_{S\in\PP\p{\ZZ_p}} \C_N\p{g_S}
    : \t{all elements of $k$ are distinct} },
  \label{eq:index_values}
\end{align}
where each $k\in\C_N\p{g}$ has the natural decomposition
$k=\p{k_S:S\in\PP\p{\ZZ_p}}$ with $k_S\in\C_N\p{g_S}$.  In words,
$\C_N\p{g}$ consists of all the ways to choose, for each subset $S$ of
the tensors $\set{w_j}$, a set $k_S$ of $g_S$ distinct spins, with the
restriction that all $k_S$ are orthogonal: if $z\in k_S$, then
$z\notin k_R$ for $R\ne S$.  Having defined the set $\C_N\p{g}$ of
index values, we define the restriction of $k\in\C_N\p{g}$ to values
associated with $w_j$ and $O_j$:
\begin{align}
  k_j \equiv \bigcup_{S\in\PP\p{\ZZ_p} : j\in S} k_S
  = \bigcup_{R\in\PP\p{\ZZ_p\setminus\set{j}}} k_{\set{j}\cup R},
\end{align}
where the union of lists $k_S$ denotes a concatenation of those lists.
These definitions allow us to expand
\begin{align}
  w\p{g} \equiv \sum_{k\in\C_N\p{g}} \prod_{j\in\ZZ_p} w_j\p{k_j},
  &&
  O\p{g} \equiv \prod_{j\in\ZZ_p}
  O_j\p{\ell_j}~\t{for some}~\ell\in\C_N\p{g},
  \label{eq:diagram_factors}
\end{align}
where the choice of $\ell\in\C_N\p{g}$ does not matter after
projection onto permutationally symmetric manifold, and
\begin{align}
  \S\p{g} \equiv \sp{\prod_{j\in\ZZ_p} { M_j \choose g_S : j\in S }}
  \sp{\prod_{S\in\PP\p{\ZZ_p}} \p{g_S!}^\abs{S}}
  = \prod_{j\in\ZZ_p} M_j!,
  \label{eq:sym_prod_symmetry}
\end{align}
where the first product in \eqref{eq:sym_prod_symmetry} contains
multinomial coefficients
\begin{align}
  { M \choose a_1, a_2, \cdots, a_m }
  \equiv \f{M!}{a_1!a_2!\cdots a_m!}
\end{align}
that account for the number of ways to partition the indices of each
$w_j,O_j$ into sets of shared indices according to $g_S$ for all
$j\in S$, and the second product in \eqref{eq:sym_prod_symmetry}
accounts for the number of ways to permute each set of shared indices
on every tensor.  Altogether, the symmetry factors $\S\p{g}$ cancel
out with the $\sim M_j!$ prefactors in \eqref{eq:sym_prod_vec}, so
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\C_N\p{M_j}} w_j\p{k} O_j\p{k}
  \EQFS \sum_{g\in\G\p{\m M}} w\p{g} O\p{g}.
  \label{eq:sym_prod_group}
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Formalizing diagrams}
\label{sec:diagrams}

Diagrams such as \eqref{eq:venn_diagram} will play a major role in the
calculations in this section.  Here, we define diagrams as formal
objects that we can use in mathematical expressions.  Formalizing
diagrams will enable us to perform necessary calculations that are
otherwise intractable.

The simplest diagrams have $g_S$ dots (``$\bullet$'') at the
intersection of the circles $j\in S$, and are simply equal to the
associated scalar $w\p{g}$ defined in \eqref{eq:diagram_factors}, e.g.
\begin{align}
  \diagram{example_triple}
  \equiv \sum_{\substack{a,b,c\in\ZZ_N\\\p{d,e}\in\C_N\p{2}\\
      \abs{\set{a,b,c,d,e}}=5}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e}.
  \label{eq:diagram_definition}
\end{align}
Additionally, we will define diagrams in which dots may be replaced by
colored crosses (``$\bm\times$'') or colored pluses (``$\bm+$'').
Whereas a diagram with $\abs{g}$ dots defines a scalar that takes
$\O\p{N^\abs{g}}$ time to compute, diagrams with crosses and pluses
will be used to compute the same scalar with a reduced computational
cost.

In a diagram with dots, crosses, and pluses, the values of indices
associated with dots are summed as if the diagram had only dots,
treating all other indices as independent and fixed.  Crosses in a
diagram then correspond to indices that are summed over all of $\ZZ_N$
without additional constraints, e.g.
\begin{align}
  \diagram{example_triple_xx_2} \equiv
  \sum_{a,c\in\ZZ_N}
  \sum_{\substack{b\in\ZZ_N\\\p{d,e}\in\C_N\p{2}\\
      \abs{\set{b,d,e}}=3}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e}.
\end{align}
If two crosses have the same color, then the corresponding indices
must be equal:
\begin{align}
  \diagram{example_triple_xx_1} \equiv
  \sum_{c\in\ZZ_N} \sum_{\substack{a,b,d\in\ZZ_N\\\abs{\set{a,b,d}}=3}}
  w_1\p{a,b,c} w_2\p{b,d,c} w_3\p{b,c,d,c}.
\end{align}
Crosses thus correspond to indices that are entirely independent of
all other indices.  Pluses, meanwhile, correspond to indices that are
{\it maximally dependent} on the dots: ``plus indices'' are summed
{\it only} over the values of ``dot indices''.  As with crosses, the
same color on two pluses indicates that the corresponding indices are
equal, e.g.
\begin{align}
  \diagram{example_triple_pp_1}
  \equiv \sum_{\substack{a\in\ZZ_N\\\p{d,e}\in\C_N\p{2}\\
      \abs{\set{a,d,e}}=3}} \sum_{b\in\set{a,d,e}}
  w_1\p{a,b,b} w_2\p{b,d,e} w_3\p{b,b,d,e}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Decomposing diagrams to reduce computational costs}
\label{sec:costs}

The diagrams generated by a product of multi-body operators in
\eqref{eq:sym_prod_group} generally have only dots, without crosses or
pluses.  Such a diagram with $\abs{g}$ dots nominally takes
$\O\p{N^{\abs{g}}}$ time to compute.  In practice, this cost can be
prohibitive for performing numerical simulations of systems of any
appreciable size.  Here, we discuss the decomposition of the diagrams
that appear in \eqref{eq:sym_prod_group} into diagrams with reduced
computational cost.  As an example, we can decompose
\begin{align}
  \diagram{example_triple}
  = \diagram{example_triple_x_top} - \diagram{example_triple_p_top},
  \label{eq:example_decomposition}
\end{align}
where
\begin{align}
  \diagram{example_triple}
  \equiv \sum_{\substack{a,b,c\in\ZZ_N\\\p{d,e}\in\C_N\p{2}\\
      \abs{\set{a,b,c,d,e}}=5}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e},
\end{align}
\begin{align}
  \diagram{example_triple_p_top}
  \equiv \sum_{\substack{b,c\in\ZZ_N\\\p{d,e}\in\C_N\p{2}\\
      \abs{\set{b,c,d,e}}=4\\a\in\set{b,c,d,e}}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e},
\end{align}
\begin{align}
  \diagram{example_triple_x_top}
  \equiv \sum_{\substack{b,c\in\ZZ_N\\\p{d,e}\in\C_N\p{2}\\
      \abs{\set{b,c,d,e}}=4}}
  w_1\p{\times,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e},
  \label{eq:example_triple_x_top}
\end{align}
with
\begin{align}
  w_1\p{\times,b,c} \equiv \sum_{a\in\ZZ_N} w_1\p{a,b,c},
\end{align}
The decomposition in \eqref{eq:example_decomposition} essentially
breaks up a sum over $a\in\ZZ_N\setminus\set{b,c,d,e}$ into a
difference of sums over $a\in\ZZ_N$ and $a\in\set{b,c,d,e}$.  The sum
over $a\in\ZZ_N$ to compute $w_1\p{\times,b,c}$ has $\O\p{N}$ cost,
and can be performed prior to computing the diagram in
\eqref{eq:example_triple_x_top}.  The decomposition in
\eqref{eq:example_decomposition} thereby reduces the calculation of
one $\O\p{N^5}$ diagram to the calculation of two $\O\p{N^4}$
diagrams.

In general, in a diagram with only dots and crosses, any dot may be
decomposed similarly to \eqref{eq:example_decomposition} into a cross
and a plus (with a new color).  The only catch in making such a
replacement is that one must manually keep track of combinatorial
factors: if the replaced dot was in a region of the diagram with $D$
dots, then the new diagrams pick up a factor of $1/D$, e.g.
\begin{align}
  \diagram{example_double}
  = \f13 \diagram{example_double_x}
  - \f13 \diagram{example_double_p}.
\end{align}
This factor can be derived by first treating the values $\bar R$ of
dots {\it outside} the region of interest fixed, and replacing the sum
over $\C_N\p{D}\setminus\bar R$ for the dots {\it within} the region
by a sum over $\ZZ_N^D\setminus\bar R$.  This replacement picks up a
factor of $1/D!$ to account for the number of ways to permute $D$
values.  After setting aside one dot for replacement by a cross and a
plus, the sum over $\ZZ_N^{D-1}\setminus\bar R$ for the remaining dots
can be changed back to a sum over $\C_N\p{D-1}\setminus\bar R$,
picking up a factor of $\p{D-1}!$.  The overall factor picked up
throughout this procedure is therefore $\p{D-1}!/D!=1/D$.

While it is possible to derive rules for decomposing dots in a diagram
with pluses, it is easier to instead eliminate pluses from a diagram
altogether.

\vspace{5cm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Reducing computational costs}

The results in \eqref{eq:triple_multi} and \eqref{eq:triple_col}
suffer from the fact that the coefficients $A_6$, $A_4$, and $A_2$
respectively take $\O\p{N^6}$, $\O\p{N^5}$, and $\O\p{N^4}$ time to
compute in general.  Here, we reduce the computational cost of these
coefficients.

We begin with the (only) diagram for $A_6$:
\begin{align}
  \diagram{triple_6}
  &= \sum_{\substack{\p{k,\ell},\p{p,q},\p{r,s}\in\C_N\p{2}\\
      \abs{\set{k,\ell,p,q,r,s}}=6}}
  u_{k\ell} v_{pq} w_{rs} \\
  &= \f12 \sum_{k,\ell\in\ZZ_N} u_{k\ell}
  \  \f12 \sum_{p,q\in\ZZ_N\setminus\set{k,\ell}} v_{pq}
  \  \f12 \sum_{r,s\in\ZZ_N\setminus\set{k,\ell,p,q}} w_{rs},
\end{align}
where we first simplify
\begin{align}
  \sum_{r,s\in\ZZ_N\setminus\set{k,\ell,p,q}} w_{rs}
  = \col{w}_{k\ell pq}
  \equiv \col{w} - 2 \sum_{r\in\set{k,\ell,p,q}} w_r
  + \sum_{r,s\in\set{k,\ell,p,q}} w_{rs},
\end{align}
with
\begin{align}
  \col{w} \equiv \sum_{r,s} w_{rs},
  &&
  w_r \equiv \sum_{s\in\ZZ_N} w_{rs}.
\end{align}
In turn,
\begin{align}
  \sum_{p,q\in\ZZ_N\setminus\set{k,\ell}} v_{pq} \, \col{w}_{k\ell pq}
  &=
\end{align}

\vspace{5cm}
